---
title: "Marker Taxa"
author: "Michelle Hagen"
output:
    html_document:
        keep_md: yes
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: show
    pdf_document:
        fig_caption: yes
        fig_height: 9
        fig_width: 8
        number_sections: yes
        toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Creating plots displaying **enrichment of taxa** towards **Drought** or
**Control** conditions for ANCOMBC, ALDEx2, DESeq2 and the SHAP value
enrichment results combined with the corresponding **significance** or
**importance** per taxon for each rank (**phylum** to **genus**).

```{r load libraries, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse); print(paste0("tidyverse, version: ", packageVersion("tidyverse")))
library(phyloseq); print(paste0("phyloseq, version: ", packageVersion("phyloseq")))
library(egg); print(paste0("egg, version: ", packageVersion("egg")))
```

## Calculate SHAP enrichment

The contribution of each taxon towards prediction of drought stress has been calculated for each fold of the cv in the `Machine_Learning.ipynb` script.
Only if there is consistency in the enrichment pattern (4/5 or 5/5 consistency), the enrichment trend is used for the taxon.

```{r threshold}
# enrichment information of 4/5 or 5/5 folds needs to be matching for a final trend
threshold <-  4
```

```{r load marker}
load("data/Phylum_marker.RData")
load("data/Class_marker.RData")
load("data/Order_marker.RData")
load("data/Family_marker.RData")
load("data/Genus_marker.RData")

all_Phylum <- Phylum_marker %>% dplyr::select(Phylum) %>% distinct()
all_Class <- Class_marker %>% dplyr::select(Class) %>% distinct()
all_Order <- Order_marker %>% dplyr::select(Order) %>% distinct()
all_Family <- Family_marker %>% dplyr::select(Family) %>% distinct()
all_Genus <- Genus_marker %>% dplyr::select(Genus) %>% distinct()

Genus_marker <- Genus_marker %>% mutate(Genus = ifelse(Genus == "Burkholderia-Caballeronia-Paraburkholderia", "Burkholderia", Genus))
```

```{r load feature importances}
Phylum_SHAP <- read.table("data/SHAP_feature_importance_phylum.csv", sep = ",", header = TRUE) %>% 
  dplyr::rename(Taxon = features)

Class_SHAP <- read.table("data/SHAP_feature_importance_class.csv", sep = ",", header = TRUE) %>% 
  dplyr::rename(Taxon = features)

Order_SHAP <- read.table("data/SHAP_feature_importance_order.csv", sep = ",", header = TRUE) %>% 
  dplyr::rename(Taxon = features)

Family_SHAP <- read.table("data/SHAP_feature_importance_family.csv", sep = ",", header = TRUE) %>% 
  dplyr::rename(Taxon = features)

Genus_SHAP <- read.table("data/SHAP_feature_importance_genus.csv", sep = ",", header = TRUE) %>% 
  dplyr::rename(Taxon = features) %>% 
  mutate(Taxon = ifelse(Taxon == "Burkholderia-Caballeronia-Paraburkholderia", "Burkholderia", Taxon))
```

```{r load enrichment information}
read_and_process_csv <- function(file_path, join_data, name_column) {
  df <- read.table(file_path, sep = ",", header = TRUE) %>%
    column_to_rownames("Taxa") %>%
    mutate(countDrought = rowSums(. == "Drought")) %>%
    mutate(Consensus = case_when(
      ((threshold == 3) & (countDrought %in% 3:5)) ~ "Drought",
      ((threshold == 3) & (countDrought %in% 0:2)) ~ "Control",
      ((threshold == 4) & (countDrought %in% 4:5)) ~ "Drought",
      ((threshold == 4) & (countDrought %in% 0:1)) ~ "Control"
    )) %>%
    dplyr::select(Consensus) %>%
    dplyr::rename(enrich = Consensus) %>%
    mutate(tool = "SHAP") %>%
    rownames_to_column(name_column) %>%
    inner_join(join_data)
  
  return(df)
}

Phylum_SHAP_enrich <- read_and_process_csv("data/SHAP_enriched_phylum.csv", all_Phylum, "Phylum") %>% 
  dplyr::rename(Taxon = Phylum)

Class_SHAP_enrich <- read_and_process_csv("data/SHAP_enriched_class.csv", all_Class, "Class") %>% 
  dplyr::rename(Taxon = Class)

Order_SHAP_enrich <- read_and_process_csv("data/SHAP_enriched_order.csv", all_Order, "Order") %>% 
  dplyr::rename(Taxon = Order)

Family_SHAP_enrich <- read_and_process_csv("data/SHAP_enriched_family.csv", all_Family, "Family") %>% 
  dplyr::rename(Taxon = Family)

Genus_SHAP_enrich <- read_and_process_csv("data/SHAP_enriched_genus.csv", all_Genus, "Genus") %>%
  mutate(Genus = ifelse(Genus == "Burkholderia-Caballeronia-Paraburkholderia", "Burkholderia", Genus)) %>% 
  dplyr::rename(Taxon = Genus)

```

The top25 significant taxa of ANCOMBC are selected for plotting. (Top 15 on Phylum rank.)

```{r ANCOMBC top 25}
ANCOMBC_top25 <- function(data, tool_name, taxon_name) {
  result <- data %>%
    dplyr::filter(tool == tool_name) %>%
    dplyr::rename(Taxon = {{taxon_name}}) %>%
    mutate(log_padj = -log10(padj)) %>%
    dplyr::select(Taxon, log_padj) %>%
    dplyr::top_n(25, wt = log_padj) %>% 
    dplyr::arrange(desc(log_padj))
  return(result)
}

ANCOMBC_top25_Phylum <- ANCOMBC_top25(Phylum_marker, "ANCOMBC2", "Phylum")
ANCOMBC_top25_Class <- ANCOMBC_top25(Class_marker, "ANCOMBC2", "Class")
ANCOMBC_top25_Order <- ANCOMBC_top25(Order_marker, "ANCOMBC2", "Order")
ANCOMBC_top25_Family <- ANCOMBC_top25(Family_marker, "ANCOMBC2", "Family")
ANCOMBC_top25_Genus <- ANCOMBC_top25(Genus_marker, "ANCOMBC2", "Genus") %>%
  mutate(Taxon = ifelse(Taxon == "Burkholderia-Caballeronia-Paraburkholderia", "Burkholderia", Taxon))
```

```{r prepare data}
Phylum_data <- Phylum_marker %>% 
  dplyr::rename(Taxon = Phylum) %>% 
  bind_rows(Phylum_SHAP_enrich) %>% 
  inner_join(ANCOMBC_top25_Phylum, by = "Taxon") %>% 
  inner_join(Phylum_SHAP, by = "Taxon") %>% 
  arrange(desc(log_padj)) %>% 
  drop_na(enrich) %>% 
  mutate(tool = ifelse(tool == "ANCOMBC2", "ANCOMBC", tool))

Class_data <- Class_marker %>% 
  dplyr::rename(Taxon = Class) %>% 
  na.omit() %>% 
  bind_rows(Class_SHAP_enrich) %>% 
  inner_join(ANCOMBC_top25_Class, by = "Taxon") %>% 
  inner_join(Class_SHAP, by = "Taxon") %>% 
  arrange(desc(log_padj)) %>% 
  drop_na(enrich) %>% 
  mutate(tool = ifelse(tool == "ANCOMBC2", "ANCOMBC", tool))

Order_data <- Order_marker %>% 
  dplyr::rename(Taxon = Order) %>% 
  na.omit() %>% 
  bind_rows(Order_SHAP_enrich) %>% 
  inner_join(ANCOMBC_top25_Order, by = "Taxon") %>% 
  inner_join(Order_SHAP, by = "Taxon") %>% 
  arrange(desc(log_padj)) %>% 
  drop_na(enrich) %>% 
  mutate(tool = ifelse(tool == "ANCOMBC2", "ANCOMBC", tool))

Family_data <- Family_marker %>% 
  dplyr::rename(Taxon = Family) %>% 
  na.omit() %>% 
  bind_rows(Family_SHAP_enrich) %>% 
  inner_join(ANCOMBC_top25_Family, by = "Taxon") %>% 
  inner_join(Family_SHAP, by = "Taxon") %>% 
  arrange(desc(log_padj)) %>% 
  drop_na(enrich) %>% 
  mutate(tool = ifelse(tool == "ANCOMBC2", "ANCOMBC", tool))

Genus_data <- Genus_marker %>% 
  dplyr::rename(Taxon = Genus) %>% 
  na.omit() %>% 
  bind_rows(Genus_SHAP_enrich) %>% 
  inner_join(ANCOMBC_top25_Genus, by = "Taxon") %>% 
  inner_join(Genus_SHAP, by = "Taxon") %>% 
  arrange(desc(log_padj)) %>% 
  drop_na(enrich) %>% 
  mutate(tool = ifelse(tool == "ANCOMBC2", "ANCOMBC", tool))
```

## Create Heatmaps + Bar plots for Taxon Enrichment/Significance/Importance

```{r}
generate_heatmap_and_barplots <- function(data) {
  custom_order <- data %>%
    select(Taxon, log_padj) %>%
    distinct() %>%
    pull(Taxon)

  tool_enrich <- data %>%
    mutate(Taxon = factor(Taxon, levels = rev(custom_order)))

  sign_imp <- data %>%
    select(Taxon, log_padj, mean_shap) %>%
    distinct()

  hm <- ggplot(data = tool_enrich, aes(x = factor(tool, levels = c("ANCOMBC", "ALDEx2", "DESeq2", "SHAP")), y = Taxon, fill = enrich)) +
    geom_tile(height = 0.9, width = 0.9) +
    coord_fixed() +
    theme_minimal(base_size = 10) +
    scale_x_discrete(position = "bottom", guide = guide_axis(angle = 45)) +
    scale_fill_manual(values = c("#5a34e3", "#f54260"), na.value = "white") +
    theme(legend.position = 'bottom',
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          axis.text.x = element_text(angle = 90),
          axis.line.x = element_line(color = 'black'),
          axis.ticks.x = element_line(color = 'black'))

  bp1 <- ggplot(data = sign_imp, aes(x = log_padj, y = reorder(Taxon, log_padj))) +
    geom_bar(stat = 'identity', width = 0.9, fill = 'orange') +
    theme_minimal(base_size = 10) +
    labs(x = expression(-log[10]("p_adjust-value"))) +
    theme(axis.text.y = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(angle = 0),
          axis.ticks.y = element_blank(),
          panel.grid = element_blank(),
          axis.line.x = element_line(color = 'black'),
          axis.ticks.x = element_line(color = 'black'))

  bp2 <- ggplot(data = sign_imp, aes(x = mean_shap, y = reorder(Taxon, log_padj))) +
    geom_bar(stat = 'identity', width = 0.9, fill = 'darkgreen') +
    theme_minimal(base_size = 10) +
    labs(x = "mean |(SHAP value)|") +
    theme(axis.text.y = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(angle = 0),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          axis.line.x = element_line(color = 'black'),
          axis.ticks.x = element_line(color = 'black'))

  egg::ggarrange(hm, bp1, bp2,
                 nrow = 1,
                 widths = c(1, 3, 3))
}

```

```{r Phylum hm, fig.cap = "**Phylum Enrichment, Significance and Importance by DAA Tools and SHAP Values.** Binary heatmap showing the enrichment of the top 15 significant phyla from ANCOMBC between Control (blue) and Drought (red) groups for the three methods used for DAA (DESeq2, ANCOMBC, ALDEx2) with an alpha <0.05, and SHAP values obtained from the RFC. Corresponding bar plots comparing $-log_{10}$(p_adjust) values (orange) and mean(|SHAP values|) (green)."}
generate_heatmap_and_barplots(Phylum_data)
```

```{r Class hm, fig.cap = "**Class Enrichment, Significance and Importance by DAA Tools and SHAP Values.** Binary heatmap showing the enrichment of the top 25 significant classes from ANCOMBC between Control (blue) and Drought (red) groups for the three methods used for DAA (DESeq2, ANCOMBC, ALDEx2) with an alpha <0.05, and SHAP values obtained from the RFC. Corresponding bar plots comparing $-log_{10}$(p_adjust) values (orange) and mean(|SHAP values|) (green)."}
generate_heatmap_and_barplots(Class_data)
```


```{r Order hm, fig.cap = "**Order Enrichment, Significance and Importance by DAA Tools and SHAP Values.** Binary heatmap showing the enrichment of the top 25 significant orders from ANCOMBC between Control (blue) and Drought (red) groups for the three methods used for DAA (DESeq2, ANCOMBC, ALDEx2) with an alpha <0.05, and SHAP values obtained from the RFC. Corresponding bar plots comparing $-log_{10}$(p_adjust) values (orange) and mean(|SHAP values|) (green)."}
generate_heatmap_and_barplots(Order_data)
```

```{r Family hm, fig.cap = "**Family Enrichment, Significance and Importance by DAA Tools and SHAP Values.** Binary heatmap showing the enrichment of the top 25 significant families from ANCOMBC between Control (blue) and Drought (red) groups for the three methods used for DAA (DESeq2, ANCOMBC, ALDEx2) with an alpha <0.05, and SHAP values obtained from the RFC. Corresponding bar plots comparing $-log_{10}$(p_adjust) values (orange) and mean(|SHAP values|) (green)."}
generate_heatmap_and_barplots(Family_data)
```

```{r Genus hm, fig.cap = "**Genera Enrichment, Significance and Importance by DAA Tools and SHAP Values.** Binary heatmap showing the enrichment of the top 25 significant genera from ANCOMBC between Control (blue) and Drought (red) groups for the three methods used for DAA (DESeq2, ANCOMBC, ALDEx2) with an alpha <0.05, and SHAP values obtained from the RFC. Corresponding bar plots comparing $-log_{10}$(p_adjust) values (orange) and mean(|SHAP values|) (green)."}
generate_heatmap_and_barplots(Genus_data)
```
